---
title: "Statistical Methods And Data Analysis In Developmental Psychology"
subtitle: "Course Preparation"
author: "Filippo Gambarota"
date: "Updated on `r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R

The course materials are created using R (4.3.0). To organize the materials I used the **R Projects** available with R Studio. I highly suggest you to use R Projects that significantly improves your workflow. A brief tutorial is available here [https://r4ds.had.co.nz/workflow-projects.html](https://r4ds.had.co.nz/workflow-projects.html).

## R Materials

All the slides and extra materials are available on Moodle. During the course and for the exercises we will use some **custom functions** that I wrote for the course. To download and use the function you can download the `utils.R` file from Moodle or from this [link](https://raw.githubusercontent.com/stat-teaching/stat-dev-psychology-2023/master/R/utils.R)(just `right-click > save`)

## R Packages

We will use, directly or indirectly (for custom functions), several R packages. You can easily install all packages using:

```{r, eval = FALSE}
pksg <- c("effects", "tidyverse", "MASS", 
          "broom", "reshape2", "ggeffects",
          "performance", "see", "car", "devtools")
install.packages(pkgs)
```

## Pipes

Sometimes in my code you will see a symbol like this `|>`, this is called **pipe**. Sometimes it is also written as `%>%` that is a different pipe coming from the `magrittr` package. The pipe is a very simple way to write R code when you need to apply multiple functions in succession. Practically, the pipe apply a function to an element, for example:

```{r, eval = FALSE}
# these two are the same
mean(x)

# to x apply the mean function
x |> mean()
```

Beyond this silly example, when we need to use multiple nested functions the pipe makes the code more readable:

```{r}
x <- runif(10)

# without pipe
exp(min(round(x, 2)))

# with pipe
x |> 
    round(2) |> 
    min() |> 
    exp()
```

Essentially, with the pipe we concatenate multiple functions where implicitly the first argument of the function is assigned to the object before the pipe.




