---
title: "Title"
subtitle: "Subtitle"
author: "Filippo Gambarota"
date: "2022-2023"
bibliography: "https://raw.githubusercontent.com/filippogambarota/bib-database/main/references.bib"
output:
    beamer_presentation:
        incremental: false
        theme: madrid
        latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      dev = "pdf", 
                      fig.align = "center",
                      out.width = "80%",
                      fig.width = 7,
                      message = FALSE, 
                      warning = FALSE)
```

```{r packages}
library(tidyverse)
library(distributional)
library(ggdist)
library(patchwork)
library(kableExtra)
devtools::load_all()
```

# Sezione 1

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more [@Zhang2012-br] details on using R Markdown see [@Williams2004-lp] <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. [@Wierzchon2016-fk]

# Sezione 2

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure, fig.cap = "ciao", out.width="50%"}
plot(pressure)
```

## References
