---
title: "Title"
subtitle: "Subtitle"
author: "Filippo Gambarota"
date: "2022-2023"
university: "@University of Padova"
email: "filippo.gambarota@unipd.it"
github: "github.com/filippogambarota"
bibliography: "https://raw.githubusercontent.com/filippogambarota/bib-database/main/references.bib"
csl: "https://raw.githubusercontent.com/citation-style-language/styles/master/apa-6th-edition.csl"
output:
  beamer_presentation:
    template: !expr here::here('template', 'slides', 'slides.tex')
    keep_tex: true
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, dev = "pdf")
```

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure, fig.cap = "ciao", out.width="50%"}
plot(pressure)
```
