---
title: "Statistical Methods and Data Analysis in Developmental Psychology"
subtitle: "Lab 8"
author: "Filippo Gambarota"
output: 
    html_document:
        code_folding: show
        toc: true
        toc_float: true
        code_download: true
date: "Updated on `r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      dev = "svg")
```

```{r packages, message=FALSE, warning=FALSE}
devtools::load_all() # if using the rproject dowloaded from the slides
# source("utils-glm.R") # if using a standard setup
library(here)
library(tidyr) # for data manipulation
library(dplyr) # for data manipulation
library(ggplot2) # plotting
library(performance) # diagnostic
library(car) # general utilities
library(MuMIn) # model selection
```

```{r options, include = FALSE}
theme_set(theme_minimal(base_size = 15))
```

```{r script, echo=FALSE}
## Link in Github repo
downloadthis::download_link(
  link = download_link("https://github.com/stat-teaching/SMDA-2023/blob/master/labs/lab8.R"),
  button_label = "Download the R script",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)
```


# Overview^[The script has been adapted from Prof. Paolo Girardi (A.Y. 2021/2022)]

We are gonna work with the `admission.csv` dataset containing $n = 400$ students for the admission to the UCLA University. A researcher is interested in how variables, such as `gre` (Graduate Record Exam scores), `gpa` (GPA), `rank` (prestige of the undergraduate institution) have an influence for the admission into graduate school. The response variable, admit/don’t admit, is a binary variable.

1. Importing data and check
2. Exploratory data analysis
3. Model fitting with `glm()`
4. Model diagnostic
5. Interpreting parameters
6. Model selection

# 1. Importing data

We need to set the **working directory** on the root of the course folder using `set.wd()`. Using R Projects is just necessary to open the `.RProj` file and the working directory will be automatically correctly selected.

```{r}
# reading data
admission <- read.csv(here("data", "admission.csv"), header=TRUE)

# first rows
head(admission)

# check dataset structure
str(admission)

# summary statistics
summary(admission)
```

It is very important that each variable is correctly interpreted by R:

- `admit` is a binary variable stored as integer (0 and 1)
- `gre` is a numerical variable stored as integer
- `gpa` is a numerical variables stored as double precision number
- `rank` is a numerical variables stored as integer

We could change the type of `rank` to factor because we are going to use it as a categorical (maybe ordinal) variable.

```{r}
admission$rankc <- factor(admission$rank, levels = 1:4, labels = 1:4)
```

# 2. Exploratory data analysis

We can plot the univariate distribution of each variable:

```{r}
# gre and gpa
admission |> 
    select(gre, gpa) |> 
    pivot_longer(1:2) |> 
    ggplot(aes(x = value)) +
    geom_histogram(col = "black",
                   fill = "lightblue") +
    facet_wrap(~name, scales = "free")
```

```{r}
admission |> 
    ggplot(aes(x = rank)) +
    geom_bar()
```

```{r}
admission |> 
    ggplot(aes(x = admit)) +
    geom_bar()
```

Then we can cut the `gpa` and `gre` variabiles into categories and plot the admissions for each bin (i.e., a contingency table):

```{r}
admission$gpa_c <- cut(admission$gpa, seq(4, 6, 0.2), labels = FALSE)
admission$gre_c <- cut(admission$gre, seq(260, 960, 50), labels=FALSE)
```

```{r}
# admission ~ gpa
admission |> 
    ggplot(aes(x = gpa_c, fill = factor(admit))) +
    geom_bar(position = position_dodge()) +
    labs(fill = "Admission") +
    theme(legend.position = "bottom")

# admission ~ gre
admission |> 
    ggplot(aes(x = gre_c, fill = factor(admit))) +
    geom_bar(position = position_dodge()) +
    labs(fill = "Admission") +
    theme(legend.position = "bottom")
```

Given that the number of admitted is lower than the number of non admitted, we can have a look at the proportion of admission for each bin:

```{r}
admission |> 
    group_by(gpa_c) |> 
    summarise(admit = mean(admit),
              non_admit = 1 - admit) |> 
    pivot_longer(2:3) |> 
    ggplot(aes(x = factor(gpa_c), y = value, fill = name)) +
    geom_col() +
    labs(fill = "Admission",
         y = "%",
         x = "gpa") +
    theme(legend.position = "bottom")

admission |> 
    group_by(gre_c) |> 
    summarise(admit = mean(admit),
              non_admit = 1 - admit) |> 
    pivot_longer(2:3) |> 
    ggplot(aes(x = factor(gre_c), y = value, fill = name)) +
    geom_col() +
    labs(fill = "Admission",
         y = "%",
         x = "gpa") +
    theme(legend.position = "bottom")
```

Finally we can have a look at the admissions as a function of the rank of the undergrad institution:

```{r}
# margin = 2 means that each colum will sum to 1
prop.table(table(admission$admit, admission$rank), margin = 2)
```

Clearly as the rank of the institute decrease (from 1 to 4) also the proportions of admissions decrease.

# 3. Model fitting with `glm()`

Now we ca fit the model using `glm()`. Let's start by fitting a *null* model with no predictors. We choose a binomial `glm` with a **logit** link function.

```{r}
fit0 <- glm(admit ~ 1, data = admission, family = binomial(link = "logit"))
summary(fit0)
```

Then we can fit the full model by putting all predictors:

```{r}
fit1 <- glm(admit ~ gre + gpa + rankc, family = binomial(link = "logit"), data = admission)
summary(fit1)
```

# 4. Model diagnostic

Firstly we can have a look to the `residual ~ fitted` plot:

```{r}
plot_resid(fit1, type = "pearson")
```

Given that the `admit` is a binary variables and we are using a bernoulli model we can use the **binned residuals** to have a better idea:

```{r}
binres <- data.frame(performance::binned_residuals(fit1, n_bins = 20))

binres |> 
    ggplot(aes(x = xbar, y = ybar)) +
    geom_point() +
    geom_line(aes(x = xbar, y = 2*se)) +
    geom_line(aes(x = xbar, y = -2*se)) +
    ylim(c(-0.5,0.5)) +
    xlab("Binned fitted(fit)") +
    ylab("Binned residuals(fit)")
```

Then we can check each predictors as a function of residuals:

```{r, fig.width=10, fig.height=10}
residualPlots(fit1, tests = FALSE)
```

Then we can check for influential observations:

```{r}
infl <- infl_measure(fit1)
head(infl)
```

Plotting using `car`

```{r}
car::influenceIndexPlot(fit1, vars = c("Studentized", "hat", "Cook"))
```

Plotting also the dfbeta:

```{r, fig.width=10, fig.height=10}
dfbeta_plot(fit1)
```

Check if there are observations with high standardized (studentized) residuals:

```{r}
outlierTest(fit1) # Testing outliers
```

For potentially influential observations we could fir a model subtracting that specific observation and compare coefficients. This is similar to the dfbeta metric that suggest no influential observations on model parameters.

```{r}
# Is 198 really influential?
fit1_no198 <- update(fit1, subset=-c(198))
compareCoefs(fit1, fit1_no198)
```

# 5. Interpreting parameters

Firstly, we can extract model parameters, taking the exponential to interpret them as odds ratios:

```{r}
broom::tidy(fit1, exponentiate = TRUE, conf.int = TRUE)
```

We can interpret these parameters as: for a unit increase in the `x`, the odds of being accepted in grad school increase by `exp(beta)`. If we multiply the `exp(beta)*100` we obtain the expected increase in percentage. Given that we have multiple parameters, when we intepret a specific parameter we are controlling for other parameters.

```{r}
broom::tidy(fit1, exponentiate = TRUE, conf.int = TRUE) |>
    slice(-1) |> 
    mutate(estperc = estimate * 100)
```

To better interpret the parameters we need to make sure that the scale is meaningful. For example, the `gre` effect seems to be very small but statistically significant. The reason is that a unit increase in `gre` is very small. We could for example rescale the variable dividing for a constant term:

```{r}
par(mfrow = c(1,2))
hist(admission$gre)
hist(admission$gre/100)
```

Let's try fitting the model with the new variable:

```{r}
admission$gre100 <- admission$gre/100
fit2 <- glm(admit ~ gre100 + gpa + rankc, family = binomial(link = "logit"), data = admission)

summary(fit2)
```

```{r}
broom::tidy(fit2, exponentiate = TRUE, conf.int = TRUE) |>
    slice(-1) |> 
    mutate(estperc = estimate * 100)
```

Now the `gre` effect is more meaningful. Notice how the overall model fitting is not changed togheter with other parameters. We are only rescaling variables.

Generally we can plot the effects for a better overview of the model:

```{r}
plot(effects::allEffects(fit1))
```

To interpret the parameters in probability terms we could use the divide by 4 rule that express the maximum slope (i.e., the maximum probability increase):

```{r}
coef(fit2)[-1]/4
```

Similarly we can compute the marginal effects for each variable that represents the average slope:

```{r}
margins::margins(fit2) |> summary()
```

Beyond the model coefficients, we could use a likelihood ratio test. Let's start by comparing the null model with the current model. We hope that our variables combinations are doing a better job compared to a null model:

```{r}
anova(fit0, fit1, test = "LRT")
```

As expected from model summary and the deviance reduction, the variables are useful to predict the probability of admission. How useful? we could use some $R^2$-like measures:

```{r}
performance::r2_tjur(fit1)
```

Despite useful, the model has a low $R^2$. Furthermore the correct classification rate is higher than the chance level but relatively low:

```{r}
1 - error_rate(fit1)
```

# 7. Model selection

We could try a model comparison starting from the null model and finishing to the overall model:

```{r}
fit2 <- update(fit2, na.action = na.fail) # required for mumin
dredge(fit2)
```

The model selection table suggest that the full model is the most appropriate, at least considering the AIC.
